---
title: "Biostat 203B Homework 2"
subtitle: Due Feb 9 @ 11:59PM
author: "Ziheng Zhang_606300061"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: false
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
knitr:
  opts_chunk: 
    cache: false    
    echo: true
    fig.align: 'center'
    fig.width: 6
    fig.height: 4
    message: FALSE
execute:
  eval: false    
---

Display machine information for reproducibility:
```{r,eval=TRUE}
sessionInfo()
```

Load necessary libraries (you can add more as needed).
```{r setup,eval=TRUE}
library(arrow)
library(data.table)
library(memuse)
library(pryr)
library(R.utils)
library(tidyverse)
```

Display memory information of your computer
```{r,eval=TRUE}
memuse::Sys.meminfo()
```

In this exercise, we explore various tools for ingesting the [MIMIC-IV](https://mimic.mit.edu/docs/iv/) data introduced in [homework 1](https://ucla-biostat-203b.github.io/2024winter/hw/hw1/hw1.html).

Display the contents of MIMIC `hosp` and `icu` data folders:

```{bash,eval=TRUE}
ls -l ~/mimic/hosp/
```

```{bash,eval=TRUE}
ls -l ~/mimic/icu/
```

## Q1. `read.csv` (base R) vs `read_csv` (tidyverse) vs `fread` (data.table)

### Q1.1 Speed, memory, and data types

There are quite a few utilities in R for reading plain text data files. Let us test the speed of reading a moderate sized compressed csv file, `admissions.csv.gz`, by three functions: `read.csv` in base R, `read_csv` in tidyverse, and `fread` in the data.table package. 
```{r,eval=FALSE}
str(read.csv("~/mimic/hosp/admissions.csv.gz"))
system.time(read.csv("~/mimic/hosp/admissions.csv.gz"))
pryr::object_size(read.csv("~/mimic/hosp/admissions.csv.gz"))
```

```{r,eval=FALSE}
str(read_csv("~/mimic/hosp/admissions.csv.gz"))
system.time(read_csv("~/mimic/hosp/admissions.csv.gz"))
pryr::object_size(read_csv("~/mimic/hosp/admissions.csv.gz"))
```

```{r,eval=FALSE}
str(fread("~/mimic/hosp/admissions.csv.gz"))
system.time(fread("~/mimic/hosp/admissions.csv.gz"))
pryr::object_size(fread("~/mimic/hosp/admissions.csv.gz"))
```

Which function is fastest? Is there difference in the (default) parsed data types? How much memory does each resultant dataframe or tibble use? (Hint: `system.time` measures run times; `pryr::object_size` measures memory usage.) 

**Answer:** `read.csv` uses 4.566 seconds and the memory usage is 158.71 MB. `read_csv` uses 1.993 seconds and the memory usage is 55.31 MB.  `fread` uses 0.557 seconds and the memory usage is 50.13 MB. So `fread` is the fastest. `read_csv` is the slowest. `read_csv` and `fread` use less memory than `read.csv`.  

The default parsed data types are different.

- For variables `admission_type`, `admission_location`, `discharge_location`, `insurance`, `language`, `religion`, `marital_status` and `race`, all three have the same data types, `character`.

- For variables `subject_id`, `hadm_id` and `hospital_expire_flag`, `read.csv` and `fread` use `integer` while `read_csv` uses `double`.

- For variables `admittime`, `dischtime`, `deathtime`, , `edregtime` and `edouttime`, `read.csv` uses `character` while `read_csv` and `fread` use `POSIXct`, a data type used to represent date and time.



### Q1.2 User-supplied data types

Re-ingest `admissions.csv.gz` by indicating appropriate column data types in `read_csv`. Does the run time change? How much memory does the result tibble use? (Hint: `col_types` argument in `read_csv`.)
```{r,eval=FALSE}
col_type <- cols(
  "admission_type" = "f",
  "admission_location" = "f",
  "discharge_location" = "f",
  "insurance" = "f",
  "language" = "f",
  "marital_status" = "f",
  "race"="f",
  "hospital_expire_flag"="f"
)
system.time(read_csv("~/mimic/hosp/admissions.csv.gz", col_types =col_type ))
pryr::object_size(read_csv("~/mimic/hosp/admissions.csv.gz"))
```
**Answer:** We change column data types for variables `admission_type`, `admission_location`, `discharge_location`, `insurance`, `language`, `marital_status`, `race` and `hospital_expire_flag` to `factor`. The run time is 1.803 seconds. The memory usage is 55.31 MB. The run time becomes shorter and the memory usage does not change.

## Q2. Ingest big data files

Let us focus on a bigger file, `labevents.csv.gz`, which is about 125x bigger than `admissions.csv.gz`.
```{bash}
#| eval: true
ls -l ~/mimic/hosp/labevents.csv.gz
```
Display the first 10 lines of this file.
```{bash}
#| eval: true
zcat < ~/mimic/hosp/labevents.csv.gz | head -10
```

### Q2.1 Ingest `labevents.csv.gz` by `read_csv`

Try to ingest `labevents.csv.gz` using `read_csv`. What happens? If it takes more than 5 minutes on your computer, then abort the program and report your findings. 
```{r,eval=FALSE}
read_csv("~/mimic/hosp/labevents.csv.gz")
```
**Answer:** It takes more than 5 minutes on my computer. So I abort the program.

### Q2.2 Ingest selected columns of `labevents.csv.gz` by `read_csv`

Try to ingest only columns `subject_id`, `itemid`, `charttime`, and `valuenum` in `labevents.csv.gz` using `read_csv`.  Does this solve the ingestion issue? (Hint: `col_select` argument in `read_csv`.)
```{r,eval=FALSE}
read_csv("~/mimic/hosp/labevents.csv.gz", col_select = c("subject_id", "itemid", "charttime", "valuenum"))
```
**Answer:** It solves the ingestion issue. It only takes around 4 minutes on my computer.

### Q2.3 Ingest subset of `labevents.csv.gz`

Our first strategy to handle this big data file is to make a subset of the `labevents` data.  Read the [MIMIC documentation](https://mimic.mit.edu/docs/iv/modules/hosp/labevents/) for the content in data file `labevents.csv`.

In later exercises, we will only be interested in the following lab items: creatinine (50912), potassium (50971), sodium (50983), chloride (50902), bicarbonate (50882), hematocrit (51221), white blood cell count (51301), and glucose (50931) and the following columns: `subject_id`, `itemid`, `charttime`, `valuenum`. Write a Bash command to extract these columns and rows from `labevents.csv.gz` and save the result to a new file `labevents_filtered.csv.gz` in the current working directory. (Hint: use `zcat <` to pipe the output of `labevents.csv.gz` to `awk` and then to `gzip` to compress the output. To save render time, put `#| eval: false` at the beginning of this code chunk.)

Display the first 10 lines of the new file `labevents_filtered.csv.gz`. How many lines are in this new file? How long does it take `read_csv` to ingest `labevents_filtered.csv.gz`?
```{bash}
#| eval: false
zcat < ~/mimic/hosp/labevents.csv.gz | awk -F, 'BEGIN{OFS=","} {if($5==50912 || $5==50971 || $5==50983 \
|| $5==50902 || $5==50882 || $5==51221 || $5==51301 || $5==50931) \
print $2,$5,$7,$10}' | gzip > ./labevents_filtered.csv.gz
```

```{bash}
#| eval: true
zcat < ./labevents_filtered.csv.gz | head -10
zcat < ./labevents_filtered.csv.gz | wc -l
```
```{r,eval=FALSE}
system.time(read_csv("./labevents_filtered.csv.gz"))
```
**Answer:** There are 24855909 lines in this new file. It takes 30.981 seconds to ingest `labevents_filtered.csv.gz` by using `read_csv`.

### Q2.4 Ingest `labevents.csv` by Apache Arrow

Our second strategy is to use [Apache Arrow](https://arrow.apache.org/) for larger-than-memory data analytics. Unfortunately Arrow does not work with gz files directly. First decompress `labevents.csv.gz` to `labevents.csv` and put it in the current working directory. To save render time, put `#| eval: false` at the beginning of this code chunk.
```{bash}
#| eval: false
zcat < ~/mimic/hosp/labevents.csv.gz > ./labevents.csv
```


Then use [`arrow::open_dataset`](https://arrow.apache.org/docs/r/reference/open_dataset.html) to ingest `labevents.csv`, select columns, and filter `itemid` as in Q2.3. How long does the ingest+select+filter process take? Display the number of rows and the first 10 rows of the result tibble, and make sure they match those in Q2.3. (Hint: use `dplyr` verbs for selecting columns and filtering rows.)
```{r,eval=TRUE}
labevents_csv <- open_dataset("./labevents.csv", format = "csv")
filtered_labevents_csv <- labevents_csv |> 
  select(subject_id, itemid, charttime, valuenum) |> 
  filter(itemid %in% c(50912, 50971, 50983, 50902, 50882, 51221, 51301, 50931)) |> 
  collect()
```
```{r,eval=FALSE}
system.time(filtered_labevents_csv <- labevents_csv |> 
              select(subject_id, itemid, charttime, valuenum) |> 
              filter(itemid %in% c(50912, 50971, 50983, 50902, 50882, 51221, 51301, 50931)) |> 
              collect())
```

```{r,eval=TRUE}
nrow(filtered_labevents_csv)
head(filtered_labevents_csv,10)
```
**Answer:** It takes 33.742 seconds for the ingest+select+filter process. There are 24855909 rows and they match those in Q2.3.

Write a few sentences to explain what is Apache Arrow. Imagine you want to explain it to a layman in an elevator.

**Answer:** Apache Arrow is like a universal language for data that allows different software systems to talk to each other more efficiently. It's a way of organizing and sharing data in a format that's easy for computers to understand and work with, kind of like a common currency that all programs can use without needing to constantly convert or translate. This makes data processing faster and smoother, helping different software programs to work together smoothly, much like how a common language helps people from different countries communicate easily, like English.

### Q2.5 Compress `labevents.csv` to Parquet format and ingest/select/filter

Re-write the csv file `labevents.csv` in the binary Parquet format (Hint: [`arrow::write_dataset`](https://arrow.apache.org/docs/r/reference/write_dataset.html).) How large is the Parquet file(s)? How long does the ingest+select+filter process of the Parquet file(s) take? Display the number of rows and the first 10 rows of the result tibble and make sure they match those in Q2.3. (Hint: use `dplyr` verbs for selecting columns and filtering rows.)
```{r,eval=TRUE}
labevents_csv <- open_dataset("./labevents.csv", format = "csv")
write_dataset(labevents_csv, format = "parquet", path = "./")
```
**Answer:** The Parquet file is 1.9G.

```{r,eval=TRUE}
file.rename("./part-0.parquet","./labevents.parquet")
```

``` {r,eval=TRUE}
labevents_parquet <- open_dataset("./labevents.parquet", format = "parquet")
filtered_labevents_parquet <- labevents_parquet |> 
  select(subject_id, itemid, charttime, valuenum) |> 
  filter(itemid %in% c(50912, 50971, 50983, 50902, 50882, 51221, 51301, 50931)) |> 
  collect()
```

```{r,eval=FALSE}
system.time(filtered_labevents_parquet <- labevents_parquet |> 
              select(subject_id, itemid, charttime, valuenum) |> 
              filter(itemid %in% c(50912, 50971, 50983, 50902, 50882, 51221, 51301, 50931)) |> 
              collect())
```

```{r,eval=TRUE}
nrow(filtered_labevents_parquet)
head(filtered_labevents_parquet,10)
```
**Answer:** It takes 16.456 seconds for the ingest+select+filter process, which is almost half of the time of the csv file. There are 24855909 rows and they match those in Q2.3.

Write a few sentences to explain what is the Parquet format. Imagine you want to explain it to a layman in an elevator.
**Answer:** Parquet is like a special way of organizing the large data so that it can be stored on many computers, kind of like splitting the table into smaller pieces and putting each piece on a different shelf. This makes it really fast to read and analyze the data, because instead of looking through the whole table at once, you can just grab the pieces you need from each shelf. So, Parquet helps big data stay organized and easy to work with, even when it's spread out across many computers. 

### Q2.6 DuckDB

Ingest the Parquet file, convert it to a DuckDB table by [`arrow::to_duckdb`](https://arrow.apache.org/docs/r/reference/to_duckdb.html), select columns, and filter rows as in Q2.5. How long does the ingest+convert+select+filter process take? Display the number of rows and the first 10 rows of the result tibble and make sure they match those in Q2.3. (Hint: use `dplyr` verbs for selecting columns and filtering rows.)
```{r,eval=TRUE}
labevents_parquet <- open_dataset("./labevents.parquet", format = "parquet")
labevents_duckdb <- to_duckdb(labevents_parquet,table = "labevents")
filtered_labevents_duckdb <- labevents_duckdb |> 
  select(subject_id, itemid, charttime, valuenum) |> 
  filter(itemid %in% c(50912, 50971, 50983, 50902, 50882, 51221, 51301, 50931)) |> 
  collect()
```

```{r,eval=FALSE}
system.time(filtered_labevents_duckdb <- labevents_duckdb |> 
              select(subject_id, itemid, charttime, valuenum) |> 
              filter(itemid %in% c(50912, 50971, 50983, 50902, 50882, 51221, 51301, 50931)) |> 
              collect())
```

```{r,eval=TRUE}
nrow(filtered_labevents_duckdb)
head(filtered_labevents_duckdb,10)
```
**Answer:** It takes 19.658 seconds for the ingest+convert+select+filter process, which is faster than the csv and almost same as the parquet file. There are 24855909 rows and they match those in Q2.3.

Write a few sentences to explain what is DuckDB. Imagine you want to explain it to a layman in an elevator.
**Answer:** DuckDB is like a super-fast and efficient way to store and work with data. It's kind of like having a super-powered calculator in your pocket that can handle huge amounts of data really quickly. DuckDB is designed to be really flexible, so you can use it for lots of different things, like analyzing data, running queries, or even building apps. It's like having a Swiss Army knife for data, making it easy to crunch numbers and find insights without having to wait around for ages. 

```{bash}
#| eval: false
rm labevents.csv
```

## Q3. Ingest and filter `chartevents.csv.gz`

[`chartevents.csv.gz`](https://mimic.mit.edu/docs/iv/modules/icu/chartevents/) contains all the charted data available for a patient. During their ICU stay, the primary repository of a patient’s information is their electronic chart. The `itemid` variable indicates a single measurement type in the database. The `value` variable is the value measured for `itemid`. The first 10 lines of `chartevents.csv.gz` are
```{bash}
#| eval: true
zcat < ~/mimic/icu/chartevents.csv.gz | head -10
```
[`d_items.csv.gz`](https://mimic.mit.edu/docs/iv/modules/icu/d_items/) is the dictionary for the `itemid` in `chartevents.csv.gz`.
```{bash}
#| eval: true
zcat < ~/mimic/icu/d_items.csv.gz | head -10
```
In later exercises, we are interested in the vitals for ICU patients: heart rate (220045), mean non-invasive blood pressure (220181), systolic non-invasive blood pressure (220179), body temperature in Fahrenheit (223761), and respiratory rate (220210). Retrieve a subset of `chartevents.csv.gz` only containing these items, using the favorite method you learnt in Q2. 

Document the steps and show code. Display the number of rows and the first 10 rows of the result tibble.

**Answer:** I used the same method as Q2.3 to ingest and filter the `chartevents.csv.gz` file.  
First step is to write a Bash command to extract these rows from `chartevents.csv.gz` and save the result to a new file `chartevents_filtered.csv.gz` in the current working directory.
```{bash}
#| eval: false
zcat < ~/mimic/icu/chartevents.csv.gz | awk -F, 'BEGIN{OFS=","} {if ($7 == 220045 || $7 == 220181 || $7 == 220179 \
|| $7 == 223761 || $7 == 220210) print}' | gzip > ./chartevents_filtered.csv.gz
```
Second step is to use bash command to display the number of rows and the first 10 rows of the new file.
```{bash}
#| eval: true
zcat < ./chartevents_filtered.csv.gz | head -10
zcat < ./chartevents_filtered.csv.gz | wc -l
```
**Answer:** There are 22502319 rows.






