---
title: "Biostat 203B Homework 5"
subtitle: Due Mar 22 @ 11:59PM
author: "Ziheng Zhang_606300061"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: false
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
execute:
  cache: true
---

## Predicting ICU duration

Using the ICU cohort `mimiciv_icu_cohort.rds` you built in Homework 4, develop at least three machine learning approaches (logistic regression with enet regularization, random forest, boosting, SVM, MLP, etc) plus a model stacking approach for predicting whether a patient's ICU stay will be longer than 2 days. You should use the `los_long` variable as the outcome. You algorithms can use patient demographic information (gender, age at ICU `intime`, marital status, race), ICU admission information (first care unit), the last lab measurements before the ICU stay, and first vital measurements during ICU stay as features. You are welcome to use any feature engineering techniques you think are appropriate; but make sure to not use features that are not available at an ICU stay's `intime`. For instance, `last_careunit` cannot be used in your algorithms. 

1. Data preprocessing and feature engineering.
```{r, eval=TRUE}
library(tidyverse)
library(tidymodels)
library(gtsummary)
library(xgboost)
library(keras)
library(vip)
library(stacks)
```

```{r, eval=TRUE}
mimiciv_icu_cohort <- readRDS("mimic_icu_cohort.rds") |>
  select(subject_id, hadm_id, stay_id, first_careunit, gender, race, 
         marital_status, age_intime, bicarbonate, chloride, creatinine, 
         glucose, potassium, sodium, hematocrit, wbc, heart_rate, 
         non_invasive_blood_pressure_systolic, 
         non_invasive_blood_pressure_diastolic, respiratory_rate, 
         temperature_fahrenheit, los_long) |>
  mutate(race = as.character(race), 
         first_careunit = as.character(first_careunit), 
         los_long = as.factor(los_long))

mimiciv_icu_cohort |> tbl_summary(by = los_long, include = -c(subject_id, 
                                                              hadm_id, 
                                                              stay_id))
```

2. Partition data into 50% training set and 50% test set. Stratify partitioning according to `los_long`. For grading purpose, sort the data by `subject_id`, `hadm_id`, and `stay_id` and use the seed `203` for the initial data split. Below is the sample code.
```{r, eval=TRUE}
set.seed(203)

# sort
mimiciv_icu_cohort <- mimiciv_icu_cohort |>
  arrange(subject_id, hadm_id, stay_id)

data_split <- initial_split(
  mimiciv_icu_cohort, 
  # stratify by los_long
  strata = "los_long", 
  prop = 0.5
  )
```

```{r, eval=TRUE}
mimic_train <- training(data_split) |>
  select(-subject_id, -hadm_id, -stay_id)
head(mimic_train)
dim(mimic_train)

mimic_test <- testing(data_split) |>
  select(-subject_id, -hadm_id, -stay_id)
head(mimic_test)
dim(mimic_test)
```
3. Train and tune the models using the training set.

**Interpretation:** There are missing values in `marital status`, all lab measurements and first vital measurements. Missing proportion is not high but there are some outliers, so we will use simple median imputation for all numerical features and mode 
imputation for `marital status`. **First we use `logstic regression` and the workflow is as follows:**
```{r, eval=TRUE}
# set recipe
logit_recipe <- recipe(los_long ~ ., data = mimic_train) |>
  
  # impute missing values with mean
  step_impute_median(bicarbonate, chloride, creatinine, glucose, potassium, 
                       sodium, hematocrit, wbc, heart_rate, 
                       non_invasive_blood_pressure_systolic, 
                       non_invasive_blood_pressure_diastolic, respiratory_rate, 
                       temperature_fahrenheit) |>
  
  # impute missing values with mode
  step_impute_mode(marital_status) |>
  
  # create traditional dummy variables
  step_dummy(all_nominal_predictors()) |>
  
  # zero-variance filter
  step_zv(all_numeric_predictors()) |> 
  
  # center and scale numeric data
  step_normalize(all_numeric_predictors()) |>
  
  print()
```

```{r, eval=TRUE}
# logistic regression
logit_mod <- 
  logistic_reg(
    penalty = tune(), 
    mixture = tune()
  ) |> 
  set_engine("glmnet", standardize = FALSE)

logit_wf <- workflow() |>
  add_recipe(logit_recipe) |>
  add_model(logit_mod)
```

```{r, eval=TRUE}
param_grid <- grid_regular(
  penalty(range = c(-5, 1)), 
  mixture(),
  levels = c(100, 5)
  )
param_grid
```

```{r, eval=TRUE}
set.seed(203)

folds <- vfold_cv(mimic_train, v = 5)

logit_fit <- logit_wf |>
  tune_grid(
    resamples = folds,
    grid = param_grid,
    metrics = metric_set(roc_auc, accuracy)
    )
```

```{r, eval=TRUE}
logit_fit |>
  # aggregate metrics from K folds
  collect_metrics() |>
  print(width = Inf) |>
  filter(.metric == "roc_auc") |>
  ggplot(mapping = aes(x = penalty, y = mean, color = factor(mixture))) +
  geom_point() +
  labs(x = "Penalty", y = "CV AUC") +
  scale_x_log10()

best_logit <- logit_fit |>
  select_best("roc_auc")

best_logit

final_wf_logit <- logit_wf |>
  finalize_workflow(best_logit)

final_fit_logit <- 
  final_wf_logit |>
  last_fit(data_split)

final_fit_logit |> 
  collect_metrics()


```
**Answer:** When `fold` = 5, the best model has `penalty`, 0.000869749 and `mixture`, 0.5. The best model has a `roc_auc` value of 0.594 and an `accuracy` of 0.572.


**Second we use `xgboost` and the workflow is as follows:**
```{r, eval=TRUE}
# set recipe
gb_recipe <- recipe(los_long ~ ., data = mimic_train) |>
  
  # impute missing values with mean
  step_impute_median(bicarbonate, chloride, creatinine, glucose, potassium, 
                       sodium, hematocrit, wbc, heart_rate, 
                       non_invasive_blood_pressure_systolic, 
                       non_invasive_blood_pressure_diastolic, respiratory_rate, 
                       temperature_fahrenheit) |>
  
  # impute missing values with mode
  step_impute_mode(marital_status) |>
  
  # create traditional dummy variables
  step_dummy(all_nominal_predictors()) |>
  
  # zero-variance filter
  step_zv(all_numeric_predictors()) |> 
  
  print()
```

```{r, eval=TRUE}
# xgboost
gb_mod <- 
  boost_tree(
    mode = "classification",
    mtry = 9,
    trees = 500, 
    tree_depth = tune(),
    learn_rate = tune()
  ) |> 
  set_engine("xgboost")

gb_wf <- workflow() |>
  add_recipe(gb_recipe) |>
  add_model(gb_mod)
```

```{r, eval=TRUE}
param_grid <- grid_regular(
  tree_depth(range = c(1L, 4L)),
  learn_rate(range = c(-3, -0.5), trans = log10_trans()),
  levels = c(4, 12)
  )

param_grid
```

```{r, eval=TRUE}
set.seed(203)

folds <- vfold_cv(mimic_train, v = 5)

gb_fit <- gb_wf |>
  tune_grid(
    resamples = folds,
    grid = param_grid,
    metrics = metric_set(roc_auc, accuracy)
    )
```

```{r, eval=TRUE}
gb_fit |>
  collect_metrics() |>
  print(width = Inf) |>
  filter(.metric == "roc_auc") |>
  ggplot(mapping = aes(x = learn_rate, y = mean, color = factor(tree_depth))) +
  geom_point() +
  labs(x = "Learning Rate", y = "CV AUC") +
  scale_x_log10()

best_gb <- gb_fit |>
  select_best("roc_auc")

best_gb

final_wf_gb <- gb_wf |>
  finalize_workflow(best_gb)

final_fit_gb <- 
  final_wf_gb |>
  last_fit(data_split)

final_fit_gb |> 
  collect_metrics()

```
**Answer:** When `fold` = 5, the best model has `learning rate`, 0.0658 and `tree depth`, 4. The best model has a `roc_auc` value of 0.636 and an `accuracy` of 0.596.



**Third we use `MLP` and the workflow is as follows:**
```{r, eval=TRUE}
# set recipe
mlp_recipe <- recipe(los_long ~ ., data = mimic_train) |>
  
  # impute missing values with mean
  step_impute_median(bicarbonate, chloride, creatinine, glucose, potassium, 
                       sodium, hematocrit, wbc, heart_rate, 
                       non_invasive_blood_pressure_systolic, 
                       non_invasive_blood_pressure_diastolic, respiratory_rate, 
                       temperature_fahrenheit) |>
  
  # impute missing values with mode
  step_impute_mode(marital_status) |>
  
  # create traditional dummy variables
  step_dummy(all_nominal_predictors()) |>
  
  # zero-variance filter
  step_zv(all_numeric_predictors()) |> 
  
  # center and scale numeric data
  step_normalize(all_numeric_predictors()) |>
  
  print()

```

```{r, eval=TRUE}
mlp_mod <- 
  mlp(
    mode = "classification",
    hidden_units = tune(),
    dropout = tune(),
    epochs = 30,
  ) |> 
  set_engine("keras", verbose = 0)

mlp_wf <- workflow() |>
  add_recipe(mlp_recipe) |>
  add_model(mlp_mod)
```

```{r, eval=TRUE}
param_grid <- grid_regular(
  hidden_units(range = c(1, 20)),
  dropout(range = c(0, 0.5)),
  levels = 5
  )
param_grid
```

```{r, eval=TRUE}
set.seed(203)

folds <- vfold_cv(mimic_train, v = 2)

mlp_fit <- mlp_wf |>
  tune_grid(
    resamples = folds,
    grid = param_grid,
    metrics = metric_set(roc_auc, accuracy)
    )
```

```{r, eval=TRUE}
mlp_fit |>
  collect_metrics() |>
  print(width = Inf) |>
  filter(.metric == "roc_auc") |>
  ggplot(mapping = aes(x = dropout, y = mean, color = factor(hidden_units))) +
  geom_point() +
  labs(x = "Dropout Rate", y = "CV AUC") +
  scale_x_log10()

best_mlp <- mlp_fit |>
  select_best("roc_auc")

best_mlp

final_wf_mlp <- mlp_wf |>
  finalize_workflow(best_mlp)

final_fit_mlp <- 
  final_wf_mlp |>
  last_fit(data_split)

final_fit_mlp |> 
  collect_metrics()

```
**Answer:** When `fold` = 2 and `epochs` = 30, the best model has `hidden_units`, 10 and `dropout`, 0. The best model has a `roc_auc` value of 0.617 and an `accuracy` of 0.585.

4. Compare model classification performance on the test set. Report both the area under ROC curve and accuracy for each machine learning algorithm and the model stacking. Interpret the results. What are the most important features in predicting long ICU stays? How do the models compare in terms of performance and interpretability?

```{r, eval=TRUE}
set.seed(203)
folds <- vfold_cv(mimic_train, v = 2)

logit_grid <- grid_regular(
  penalty(range = c(-5, 1)), 
  mixture(),
  levels = c(100, 5)
  )

gb_grid <- grid_regular(
  tree_depth(range = c(1L, 4L)),
  learn_rate(range = c(-3, -0.5), trans = log10_trans()),
  levels = c(4, 12)
  )

mlp_grid <- grid_regular(
  hidden_units(range = c(1, 20)),
  dropout(range = c(0, 0.5)),
  levels = 5
  )

logit_res <- 
  tune_grid(
    object = logit_wf, 
    resamples = folds, 
    grid = logit_grid,
    control = control_stack_grid()
  )


```

```{r, eval=TRUE}
gb_res <-
  tune_grid(
    object = gb_wf, 
    resamples = folds, 
    grid = gb_grid,
    control = control_stack_grid()
  )
```

```{r, eval=TRUE}
mlp_res <-
  tune_grid(
    object = mlp_wf, 
    resamples = folds, 
    grid = mlp_grid,
    control = control_stack_grid()
  )
```

```{r, eval=TRUE}
mimic_model_st <- 
  # initialize the stack
  stacks() |>
  # add candidate members
  add_candidates(logit_res) |>
  add_candidates(gb_res) |>
  add_candidates(mlp_res) |>
  # determine how to combine their predictions
  blend_predictions(
    penalty = 10^(-5:1),
    metrics = c("roc_auc"), 
    times = 3
    ) |>
  # fit the candidates with nonzero stacking coefficients
  fit_members()
```

```{r, eval=TRUE}
mimic_model_st

autoplot(mimic_model_st)

autoplot(mimic_model_st, type = "weights")
```

```{r, eval=TRUE}
mimic_pred <- mimic_test |>
  bind_cols(predict(mimic_model_st, ., type = "prob")) |>
  mutate(predicted = if_else(.pred_Yes > 0.5, "FALSE", "TRUE")) |>
  print(width = Inf)
```

```{r, eval=TRUE}
final_fit_logit |> 
  collect_metrics() |>
  select(.metric, .estimate)

final_fit_gb |> 
  collect_metrics() |>
  select(.metric, .estimate)

final_fit_mlp |> 
  collect_metrics() |>
  select(.metric, .estimate)

yardstick::accuracy(
  mimic_pred,
  truth = los_long,
  estimate = predicted
  )

yardstick::roc_auc(
  mimic_pred,
  truth = los_long,
  contains(".pred_Yes")
  )
```

```{r, eval=TRUE}
# feature importance

final_fit_logit |> 
  extract_fit_parsnip() |>
  vip()

final_fit_gb |> 
  extract_fit_parsnip() |>
  vip()

final_fit_mlp |> 
  extract_fit_parsnip() |>
  vip()


```






